---
title: "Analyzing eCommerce data"
author: "Tommy Wang, Sadaf Sultan, Jean-Francois Lafon, Alexandre Le Cann" 
output:
  html_document:
    css: ./AnalyticsStyles/default.css
    theme: paper
    toc: yes
    toc_float:
      collapsed: no
      smooth_scroll: yes
always_allow_html: yes
---

```{r setuplibraries, echo=FALSE, message=FALSE}
# SET UP
local_directory <- "."
suppressWarnings(source(paste(local_directory,"AnalyticsLibraries/library.R", sep="/")))

# Package options
suppressWarnings(ggthemr('fresh'))  # ggplot theme
opts_knit$set(progress=FALSE, verbose=TRUE)
opts_chunk$set(echo=FALSE, fig.align="center", fig.width=10, fig.height=6.35, results="asis")
options(knitr.kable.NA = '')

# Determine document output format, return "html" by default
getDocumentOutputFormat <- function() {
  format <- opts_knit$get('rmarkdown.pandoc.to')
  if (!is.null(format)) format else "html"
}

# Format tables for html/latex output
normalize.abs <- function(x, min=0, max=1, na.rm=FALSE) normalize(abs(x), min, max, na.rm)
iprint.df <- function(df, scale=FALSE) {
    if (class(df) != "data.frame")
      df <- as.data.frame(df)
    x <- lapply(colnames(df), function(col) {
      if (is.numeric(df[, col]))
        color_bar(rgb(238, 238, 238, max=255), normalize.abs, min=0.1, na.rm=TRUE)
      else
        formatter("span")
    })
    names(x) <- colnames(df)
    tags$div(class="formattable_container", HTML(gsub("NA", "", format_table(df, x))))
  } 

# Format plots for html/latex output
iplot.df <- function(df, x=colnames(df)[1], y="value", v="variable", type="line", xlab=NULL, ylab=NULL) {
    p <- c3(df, x=x, y=y, group=v, width="100%", height="480px")
    p <- switch(type,
      line = p %>% c3_line('spline'),
      bar  = p %>% c3_bar(bar_width=0.90)
    )
    if (!is.null(xlab)) p <- p %>% xAxis(label=xlab)
    if (!is.null(ylab)) p <- p %>% yAxis(label=ylab)
    p
  }

iplot.hist <- function(x, breaks="Sturges", xlab=NULL) {
  h <- hist(x, breaks=breaks, plot=FALSE)
  df <- data.frame(x=head(h$breaks, -1), Frequency=h$counts)
  iplot.df(df, x="x", y="Frequency", v=NULL, type="bar", xlab=xlab)
}

iplot.grid <- if (getDocumentOutputFormat() == "html") tags$div else grid.arrange

iplot.dendrogram <- function(cluster) {
  labels <- (length(cluster$labels) > 40)
  if (getDocumentOutputFormat() == "html") {
    cluster$labels <- if (!labels) NULL else cluster$labels
    margins <- list(top=10, right=0, bottom=ifelse(labels, 120, 10), left=0)
    dendroNetwork(cluster, width="100%", height="480px", fontSize=14,
                  treeOrientation="vertical", margins=margins, textRotate=90)
  } else {  # latex, etc.
    ggdendrogram(Hierarchical_Cluster, theme_dendro=FALSE, labels=labels) +
      xlab("Observations") + ylab("Height")
  }
}

```

We have gathered data from Ecommerce website sales for 3 days. The goal of our analysis is to segment our customers 

#Importing the data

First step is of course to import the data.

Here swe have two data files:  

* "Anonymized_transactions_all.csv" which compiles all transactions that happened between January 1st and February 7th. Each row of the file corresponds to the purchase of one item (when various items are bought during the same transaction, several lines are created). The different fields listed in the file are:
  + Order numer: unique identification of the order
  + SKU number: identifier of an item in the ecommerce database
  + Unit price of the item purchased
  + Price paid for the item (can be different from the item price if there are promotions for example)
  + Date of the order
  + Time of the order
  + Payment method used
  + Price paid for the whole order
  + Anonymized name of the customer ("Anonym"+several digits)
  
* "Categories_all.csv" contains a breakdown of each item into sub categories. Each line corresponds to an item that can be purchased through the ecommerce website. The different fields are:
  + SKU number: the unique identifier for each item
  + Item description: a sentence describing the item
  + Category of the item
  + Sub-category of the item
  + Sub-sub-category of the item 

```{r, echo=FALSE, tidy=TRUE}
# Please ENTER the name of the files with the data used. 
Transaction_datafile_name = "Anonymized_transactions_all.csv"
Categories_datafile_name = "Categories_all.csv"

# Please enter the minimum number below which you would like not to print - this makes the readability of the tables easier. Default values are either 10e6 (to print everything) or 0.5. Try both to see the difference.
MIN_VALUE = 0.5

# Please enter the maximum number of observations to show in the report and slides. 
# DEFAULT is 10. If the number is large the report may be slow.
max_data_report = 10
```

```{r, echo=TRUE, tidy=TRUE}
#This simply reads the CSV files and creates two variables containing text data
Transaction_data <- read.csv(Transaction_datafile_name, sep=";")
Categories_data <- read.csv(Categories_datafile_name, sep=";")



#For later use, we have to remove the items that don't have any category from the dataset
#Variable SKU_Cat will be a vector containing the categories and whose names are the SKUs present in the datafile "Categories_all.csv"
SKU_Cat <-Categories_data[,"Category"] 
names(SKU_Cat)<-Categories_data[,"SKU"]

#We have to exclude some customers, linked to fake accounts (created for debug purposes by the ecommerce website)

Customer_blacklist=c("Customer254", "Customer262", "Customer302", "Customer9869")
Transaction_blacklist=NULL

#Next, let's add to this blacklist the transactions involving SKU not present in the datafile "Categories_all.csv"
for (i in 1:nrow(Transaction_data)){
  if (!(as.character(Transaction_data[i,"SKU"]) %in% names(SKU_Cat))||(as.character(Transaction_data[i,"Anonym"]) %in% Customer_blacklist)){
    Transaction_blacklist=c(Transaction_blacklist,i)
  }
}

#we keep only the ones that are not on the blacklist
Transaction_data<-Transaction_data[-Transaction_blacklist,]


nbTransactions=nrow(Transaction_data)


```

After having imported the data, the idea is to create a new variable, that sums up the items bought by each customer. 

```{r, echo=TRUE, tidy=TRUE}
#We have to figure out at first the list of customers and of items sold
Customer_list = as.character(Transaction_data[1,"Anonym"])
ItemsSold_list = as.character(Transaction_data[1,"SKU"])
Orders_list = as.character(Transaction_data[1,"Order.Number"])
for (i in 2:nbTransactions) {
  if (!(as.character(Transaction_data[i,"Anonym"]) %in% Customer_list)){
    Customer_list<-c(Customer_list, as.character(Transaction_data[i,"Anonym"]))
  }
  if (!(as.character(Transaction_data[i,"SKU"]) %in% ItemsSold_list)){
    ItemsSold_list<-c(ItemsSold_list, as.character(Transaction_data[i,"SKU"]))
  }
    if (!(as.character(Transaction_data[i,"Order.Number"]) %in% Orders_list)){
    Orders_list<-c(Orders_list, as.character(Transaction_data[i,"Order.Number"]))
  }
}

#The number of unique customers and different items sold is then easy to compute
nbCustomers=length(Customer_list)
nbItemsSold=length(ItemsSold_list)
nbOrders=length(Orders_list)

#Then we create a matrix that we will populate with the actual transactions
Sales_Cust_Items = matrix(0*1:(nbCustomers*nbItemsSold), ncol=nbItemsSold, nrow=nbCustomers)
colnames(Sales_Cust_Items) <- ItemsSold_list
rownames(Sales_Cust_Items) <- Customer_list

for (i in 1:nbTransactions){
  itemsold=as.character(Transaction_data[i,"SKU"])
  customer=as.character(Transaction_data[i,"Anonym"])
  Sales_Cust_Items[customer,itemsold]<-Sales_Cust_Items[customer,itemsold]+1
}

#And we create a matrix that we will populate with the actual orders
Sales_Cust_Orders = matrix(0*1:(nbCustomers*nbOrders), ncol=nbOrders, nrow=nbCustomers)
colnames(Sales_Cust_Orders) <- Orders_list
rownames(Sales_Cust_Orders) <- Customer_list

for (i in 1:nbTransactions){
  order=as.character(Transaction_data[i,"Order.Number"])
  customer=as.character(Transaction_data[i,"Anonym"])
  Sales_Cust_Orders[customer,order]<-Sales_Cust_Orders[customer,order]+1
}

```

We have created 2 variables: Sales_Cust_Items and Sales_Cust_Orders

#Basic data visualizations
##Distribution of the number of items ordered per customer

A total of `r nbItemsSold` items have been sold over the period

```{r}
# Plot of the distribution of number of items purchased on the whole period by each customer

Items_sold_per_customer = as.data.frame(apply(Sales_Cust_Items,1,sum))
names(Items_sold_per_customer)<-"Number_of_Items_Purchased"

ggplot(Items_sold_per_customer, aes(Items_sold_per_customer$Number_of_Items_Purchased)) + geom_histogram(fill = '#B8274C', binwidth = 0.5) + scale_x_continuous(breaks = seq(0,15), limits = c(0, 15)) + labs(x = "Number of orders",y = "Number of items purchased")

```


```{r}
# Plot of the distribution of number of items on the whole period by each customer

Items_sold_per_customer2 = as.data.frame(table(Transaction_data$Anonym))

ggplot(Items_sold_per_customer2, aes(Items_sold_per_customer2$Freq)) + geom_histogram(fill = '#B8274C', binwidth = 0.5) + scale_x_continuous(breaks = seq(0,15), limits = c(0, 15)) + labs(x = "Number of orders")

```


##Distribution of number of orders per customer

A total of `r nbOrders` orders have been made over the period
The majority of customers have only made 1 order

```{r}
# Plot of the distribution of number of orders on the whole period by each customer

Sales_Cust_Orders[Sales_Cust_Orders > 1] <- 1

Orders_per_customer = as.data.frame(apply(Sales_Cust_Orders,1,sum))
names(Orders_per_customer)<-"Number_of_Orders"

ggplot(Orders_per_customer, aes(Orders_per_customer$Number_of_Orders)) + geom_histogram(fill = '#B8274C', binwidth = 0.5) + scale_x_continuous(breaks = seq(0,15), limits = c(0, 15)) + labs(x = "Number of orders")

```

##Top 10 customers (by number of items ordered)

Top 10 customers bought `r round(sum(Top10_customers[,2])/nbItemsSold*100,digits = 2)`% of all items sold over the period 

```{r}
# Plot of the distribution of number of items on the whole period by the top 10 customers

Items_sold_per_customer2 = as.data.frame(table(Transaction_data$Anonym))
Customers_sorted <- Items_sold_per_customer2[order(Items_sold_per_customer2[,2], decreasing=TRUE),]
Top10_customers <- Customers_sorted[1:10,]
rownames(Top10_customers) <- c(1:10)
ggplot(Top10_customers, aes(x = reorder(Var1,-Freq), y=Freq)) + geom_bar(stat = "identity") + theme(axis.text.x = element_text(angle = 45, hjust = 1)) + labs(x = "Top 10 customers", y = "Number of items ordered")

```

##Top 10 SKUs (by number of items ordered)

Top 10 SKUs represent `r round(sum(Top10_SKUs[,2])/nbItemsSold*100,digits = 2)`% of all items sold over the period

```{r}
# Plot of the distribution of number of items sold on the whole period for the Top 10 bestsellers

Items_sold_per_customer3 = as.data.frame(table(Transaction_data$SKU))
SKUs_sorted <- Items_sold_per_customer3[order(Items_sold_per_customer3[,2], decreasing=TRUE),]
Top10_SKUs <- SKUs_sorted[1:10,]
rownames(Top10_SKUs) <- c(1:10)
ggplot(Top10_SKUs, aes(x = reorder(Var1,-Freq), y=Freq)) + geom_bar(stat = "identity") + theme(axis.text.x = element_text(angle = 45, hjust = 1)) + labs(x = "Top 10 SKUs", y = "Number of items ordered")


```


#Segmentation of the customers
The goal of our analysis is to build an algorithm that would suggest other purchases to existing customers.

Our data goes from January 1st to February 7th. We will divide the dataset in two:
* First dataset regroups people who bought items in January and in February
* Second dataset regroups people who ordered more than one items during January, excluding the customers who are already in the first dataset.

We will use the second dataset to buid a segmentation of the customer based on the items they bought, and then use the second dataset to check if the segmentation is accurate enough to be predicitve: we will assign the customers from the first dataset in segments based on their purchases in January, and try then to anticipate their purchases in February.

```{r, echo=FALSE}

#The first step is to find the customers that purchased both in January and February

#The dataset Transaction_data is already ordered by date, let's find the first purchase made in February 

FirstPurchaseFeb = 1

while((as.character(Transaction_data[FirstPurchaseFeb,"Created.Date"])!="01/02/2013")&(FirstPurchaseFeb<nbTransactions+1)) {
  FirstPurchaseFeb<-FirstPurchaseFeb+1
}

Customer_Jan <- as.character(Transaction_data[1:(FirstPurchaseFeb-1), "Anonym"])
Customer_Feb <- as.character(Transaction_data[FirstPurchaseFeb:nbTransactions, "Anonym"])
Customer_JnF <- intersect(Customer_Jan, Customer_Feb)

#Finally, we have to eliminate redundant customers from CustomerJnF

Single_Customer_JnF=unique(Customer_JnF)


#The variable Single_Customer_JnF now contains a list of non-redundant customers who both bought on January and February

#Then we build the second dataset, customers who have purchased more than one item, excluding those that are already in the first dataset

Item_threshold = 2 # We need to be able to adjust the relevant nb of items purchased

#We will at first extract the relevant customers and the items they bought
Customer_segm=NULL
Items_segm=NULL

for (i in 1:nrow(Sales_Cust_Items)){
  if ((sum(Sales_Cust_Items[i,])>=Item_threshold)&!(as.character(rownames(Sales_Cust_Items)[i]) %in% Single_Customer_JnF)){
    Customer_segm=c(Customer_segm, as.character(rownames(Sales_Cust_Items)[i]))
    Item_to_add=names(Sales_Cust_Items[i,Sales_Cust_Items[i,]>0])
    Items_segm=c(Items_segm, as.character(Item_to_add))
  }
}

Items_segm=unique(Items_segm)

# Now the variable Customer_segm contains all the customers that bought several items but only in January. The variable Items_segm contains the items they bought.
# Let's build the matrix that contains the number of items they bought and store it in the variable Matrix_segm
# The rows of the matrix will be the customers and the column will be the items they bought

Matrix_segm=matrix(0*1:(length(Items_segm)*length(Customer_segm)), ncol = length(Items_segm), nrow = length(Customer_segm))

colnames(Matrix_segm)<-Items_segm
rownames(Matrix_segm)<-Customer_segm

Matrix_segm[Customer_segm,Items_segm]<-Sales_Cust_Items[Customer_segm,Items_segm]


```

##Data vizualization

We can display data like:
* Number of items sold per customer on average
* Distribution of number of items sold
* Amount spent per customer
* ...

```{r, echo=TRUE}
# Plot of the distribution of number of items purchased on the whole period by each customer
Items_sold_per_customer_category = as.data.frame(apply(Matrix_segm,1,sum))
names(Items_sold_per_customer_category)<-"Number_of_Items_Purchased"

ggplot(Items_sold_per_customer_category, aes(Items_sold_per_customer_category$Number_of_Items_Purchased)) + geom_histogram(fill = '#B8274C', binwidth = 0.5) + scale_x_continuous(breaks = seq(0,15), limits = c(0, 15)) + labs(x = "Number of orders")

```

##Analysis of the results

TO BE COMPLETED

#Use of categories instead of item name

To make our computatinos easier, we will associate each item to its category, thus drastically reducing the number of columns in the matrix.

```{r}
#Computing the correspondance matrix between SKU and Category.  
Category_list = SKU_Cat[colnames(Matrix_segm)]
Category_list<-unique(Category_list)

Category_matrix=matrix(0*1:(length(Category_list)*nrow(Matrix_segm)),nrow=nrow(Matrix_segm), ncol=length(Category_list))

rownames(Category_matrix) <- Customer_segm
colnames(Category_matrix) <- Category_list


for (i in 1:nrow(Transaction_data)) {
  Customer_tmp<-as.character(Transaction_data[i,"Anonym"])
  Item_tmp<-as.character(Transaction_data[i,"SKU"])
  Category_tmp<-as.character(SKU_Cat[Item_tmp])
   
  if (Customer_tmp %in% Customer_segm){
    Category_matrix[Customer_tmp,Category_tmp]<-Category_matrix[Customer_tmp,Category_tmp]+1
  }
}

```

```{r}
# Plot of the distribution of number of items purchased on the whole period by each customer
Items_sold_per_category = as.data.frame(apply(Category_matrix,2,sum))
names(Items_sold_per_category)<-"Number_of_Items_Purchased_Per_Cat"

ggplot(Items_sold_per_category, aes(x=reorder(rownames(Items_sold_per_category),-apply(Category_matrix,2,sum)),y=apply(Category_matrix,2,sum))) + geom_bar(stat = "identity") + theme(axis.text.x = element_text(angle = 45, hjust = 1)) + labs(x = "Categories", y = "Number of items ordered")

```

##PCA analysis

```{r}

# Here is how the `principal` function is used 
UnRotated_Results<-principal(Category_matrix, nfactors=ncol(Category_matrix), rotate="none",score=TRUE)
UnRotated_Factors<-round(UnRotated_Results$loadings,2)
UnRotated_Factors<-as.data.frame(unclass(UnRotated_Factors))
colnames(UnRotated_Factors)<-paste("Comp",1:ncol(UnRotated_Factors),sep="")

```

```{r}
# Here is how we use the `PCA` function 
Variance_Explained_Table_results<-PCA(Category_matrix, graph=FALSE)
Variance_Explained_Table<-Variance_Explained_Table_results$eig
Variance_Explained_Table_copy<-Variance_Explained_Table

rownames(Variance_Explained_Table) <- paste("Component", 1:nrow(Variance_Explained_Table), sep=" ")
colnames(Variance_Explained_Table) <- c("Eigenvalue", "Pct of explained variance", "Cumulative pct of explained variance")
```

Let's look at the **variance explained** as well as the **eigenvalues** :

```{r}
iprint.df(round(Variance_Explained_Table, 2))
```

```{r}
eigenvalues  <- Variance_Explained_Table[, "Eigenvalue"]
df           <- cbind(as.data.frame(eigenvalues), c(1:length(eigenvalues)), rep(1, length(eigenvalues)))
colnames(df) <- c("eigenvalues", "components", "abline")
iplot.df(melt(df, id="components"))

```


#Use of Sub-categories

Not every item has a sub_category in the file "Categories_all.csv", so we will define the subcategory as the concatenation of the Category and the Subcategory found in the datafile.

It means that if an item doesn't have a subcategory, it will be referenced by its category.



```{r}
#Computing the correspondance matrix between SKU and Sub-Category.  
SKU_SubCat <-paste(SKU_Cat,Categories_data[,"SubCategory"],sep="_")
names(SKU_SubCat)<-Categories_data[,"SKU"]

SubCategory_list = SKU_SubCat[colnames(Matrix_segm)]
SubCategory_list<-unique(SubCategory_list)

SubCategory_matrix=matrix(0*1:(length(SubCategory_list)*nrow(Matrix_segm)),nrow=nrow(Matrix_segm), ncol=length(SubCategory_list))

rownames(SubCategory_matrix) <- Customer_segm
colnames(SubCategory_matrix) <- SubCategory_list


for (i in 1:nrow(Transaction_data)) {
  Customer_tmp<-as.character(Transaction_data[i,"Anonym"])
  Item_tmp<-as.character(Transaction_data[i,"SKU"])
  SubCategory_tmp<-as.character(SKU_SubCat[Item_tmp])
   
  if (Customer_tmp %in% Customer_segm){
    SubCategory_matrix[Customer_tmp,SubCategory_tmp]<-SubCategory_matrix[Customer_tmp,SubCategory_tmp]+1
  }
}

```

```{r}
# Plot of the distribution of number of items purchased on the whole period by each customer
Items_sold_per_subcategory = as.data.frame(apply(SubCategory_matrix,2,sum))
names(Items_sold_per_subcategory)<-"Number_of_Items_Purchased_Per_SubCat"

ggplot(x=rownames(Items_sold_per_subcategory),
      y= apply(SubCategory_matrix,2,sum), 
      geom="hist") 

```

##PCA analysis

```{r}

# Here is how the `principal` function is used 
UnRotated_Results<-principal(SubCategory_matrix, nfactors=ncol(SubCategory_matrix), rotate="none",score=TRUE)
UnRotated_Factors<-round(UnRotated_Results$loadings,2)
UnRotated_Factors<-as.data.frame(unclass(UnRotated_Factors))
colnames(UnRotated_Factors)<-paste("Comp",1:ncol(UnRotated_Factors),sep="")

```

```{r}
# Here is how we use the `PCA` function 
Variance_Explained_Table_results<-PCA(SubCategory_matrix, graph=FALSE)
Variance_Explained_Table<-Variance_Explained_Table_results$eig
Variance_Explained_Table_copy<-Variance_Explained_Table

rownames(Variance_Explained_Table) <- paste("Component", 1:nrow(Variance_Explained_Table), sep=" ")
colnames(Variance_Explained_Table) <- c("Eigenvalue", "Pct of explained variance", "Cumulative pct of explained variance")
```

Let's look at the **variance explained** as well as the **eigenvalues** :

```{r}
iprint.df(round(Variance_Explained_Table, 2))
```

```{r}
eigenvalues  <- Variance_Explained_Table[, "Eigenvalue"]
df           <- cbind(as.data.frame(eigenvalues), c(1:length(eigenvalues)), rep(1, length(eigenvalues)))
colnames(df) <- c("eigenvalues", "components", "abline")
iplot.df(melt(df, id="components"))
```



The goal of this section is to test if the segmentation of the second dataset is able to predict the purchases from the first dataset.

----------
 CODE TO BE INSERTED here
-------------
